# Class 15: 17 11 2020 {#aula15}

Model selection and assessment - this chapter/class 
is currently lame at best, no good writting etc.

# Example 1 - deviance

A15 ME 17 11 2019.pptx
slides 22-25

# Example 2 - lagartyos GLM

A15 ME 17 11 2019.pptx
slides 26-33 

## Example 2

Use the aerogeradores example to see how AIC and significance tests on the parameters can tell you different things

Read in the data

```{r}
peol <- read.csv("extfiles/parqueseolicos.csv", sep=";")
```

Look at the data

```{r}
pairs(peol)
```

Implement multiple regression

```{r}
regmul=lm(mortalidade~.,data=peol)
summary(regmul)
```

Implement poisson GLM with all variables

```{r}
glmALL=glm(mortalidade~.,data=peol,family=poisson(link="log"))
summary(glmALL)
```

Run a few other models

```{r}
glmda<-glm(mortalidade~densidade.aero,data=peol,family=poisson(link="log"))
glma<-glm(mortalidade~altitude,data=peol,family=poisson(link="log"))
glmpa<-glm(mortalidade~proximidade.aero,data=peol,family=poisson(link="log"))
glmapa<-glm(mortalidade~altitude+proximidade.aero,data=peol,family=poisson(link="log"))
glmdaapa=glm(mortalidade~densidade.aero+altitude+proximidade.aero,data=peol,family=poisson(link="log"))
```


Check that best model by AIC inclues non significant variables

```{r}
AIC(glmda,glma,glmpa,glmapa,glmdaapa,glmALL)
```

Check best AIC model

```{r}
summary(glmdaapa)
```

Use `stepAIC`

```{r}
library(MASS)
fullmodel=glm(mortalidade~.,data=peol,family=poisson(link="log"))
best.model=stepAIC(fullmodel,direction="backward")

summary(best.model)
```

Use function `bestglm`, confirm you get the best above and also the same as from stepAIC

```{r}
library(bestglm)
bestGLM=bestglm(Xy=peol[,c(2:5,1)],family=poisson(),IC="AIC",RequireFullEnumeration=TRUE)
bestGLM$BestModel
```

Interesting, with the Gaussian response (and I used the Gaussian response on the slides in class!) one would get a different answer... explore?

```{r}
library(bestglm)
bestGLM=bestglm(Xy=peol[,c(2:5,1)],family=gaussian(),IC="AIC",RequireFullEnumeration=TRUE)
bestGLM$BestModel
```

Any way, do not forget, these were just an example dataset to explore functions stepAIC and bestglm. If ecological inferences were the obnjective, the answer to our doubts would be: find a reasonable dataset, not one with `r ncol(peol)-1` independent variables and `r nrow(peol)` data points.

Look at AICc?