# Class 16: 18 11 2020 {#aula16}

## A logistic regression example

```{r}

#
set.seed(123)
n=200
#a covariate
xs=runif(n,-20,20)
#get the mean value
ilogit=function(x){
  il=exp(x)/(1+exp(x))
return(il)
}

Ey=ilogit(2+0.4*xs)
#generate data
ys=rbinom(n,size = rep(1,n),prob = Ey)
#plot data
plot(xs,ys)


#run a glm
glmLR1=glm(ys~xs,family=binomial(link="logit"))
summary(glmLR1)

xs4pred=seq(min(xs),max(xs),length=100)
predglmLR1=predict(glmLR1,newdata = data.frame(xs=xs4pred),
type="response")

par(mfrow=c(1,1),mar=c(4,4,0.2,0.2))
plot(xs,ys)
lines(xs4pred,predglmLR1,col=3,lty=2)

summary(glmLR1)


#check model fit
par(mfrow=c(2,2),mar=c(4,4,2,0.2))
plot(glmLR1)

```

## About multicollinearity

(this is probably best way before, in the linear model stuff... any way, I presented it here as this topic started from a student question)

```{r}
#Código da aula 11

#--------------------------------------------------------
# When good models go wrong as 
# multicollinearity kicks in
#--------------------------------------------------------
library(MASS)
set.seed(1234)
n=100
means <- c(2,4,6,8,10,12)
ncovs=(36-6)/2
covs<- rnorm(ncovs,mean=10,sd=2)
varcovars=matrix(NA,6,6)
varcovars[lower.tri(varcovars)]=covs
varcovars=t(varcovars)
varcovars[lower.tri(varcovars)]=covs
diag(varcovars)=means
varcovars=t(varcovars) %*% varcovars
indvars <- mvrnorm(n = n, mu=means, Sigma=varcovars)

# we can see that we have high correlations 
#across the board in explanatory variables
indvars=as.data.frame(indvars)
names(indvars)=paste0("X",1:6)
head(indvars)
round(cor(indvars),2)

ys <-510+4*indvars$X1+rnorm(n,mean=0,sd=200)
par(mfrow=c(1,1),mar=c(4,4,0.5,0.5))
plot(ys~indvars$X1)
lmX1 <- lm(ys~indvars$X1)
abline(lmX1)
summary(lmX1)

plot(ys~indvars$X2)
lmX2 <- lm(ys~indvars$X2)
abline(lmX2)
summary(lmX2)

#one error type I + 1 error type 2
lmX1X2 <- lm(ys~indvars$X1+indvars$X2)
summary(lmX1X2)


lmX3 <- lm(ys~indvars$X3)
summary(lmX3)
lmX1X3 <- lm(ys~indvars$X1+indvars$X3)
summary(lmX1X3)

lmX4 <- lm(ys~indvars$X4)
summary(lmX4)
lmX1X4 <- lm(ys~indvars$X1+indvars$X4)
summary(lmX1X4)

lmX5 <- lm(ys~indvars$X5)
summary(lmX5)
lmX1X5 <- lm(ys~indvars$X1+indvars$X5)
summary(lmX1X5)


lmX6 <- lm(ys~indvars$X6)
summary(lmX6)
lmX1X6 <- lm(ys~indvars$X1+indvars$X6)
summary(lmX1X6)

AIC(lmX1,lmX2,lmX3,lmX4,lmX5,lmX6,lmX1X2,lmX1X3,lmX1X4,lmX1X5,lmX1X6)

```

The fundamental idea is that most of these models are equally good for prediction. But they might give you the complete wrong picture for explanation. Never confuse the two. Optimal modelling strategies and choices might require know what is more important.
